---
title: "lab12 population scale analysis"
author: "Longmei Zhang A17012012"
format: pdf
---

```{r}
tb <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

```

```{r}
table(tb$Genotype..forward.strand.) * 100 / nrow(tb)

```

Population Scale Analysis
One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3
expression.

read the data
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
nrow(expr)
```
```{r}
table(expr$geno)
```
```{r}
ag <- expr$geno == "A/G"
ag_expr = expr[ag,]
summary(ag_expr$exp)

gg <- expr$geno == "G/G"
gg_expr = expr[gg,]
summary(gg_expr$exp)

aa <- expr$geno == "A/A"
aa_expr = expr[aa,]
summary(aa_expr$exp)
```

> Q13 Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

There are 462 samples in total, 108 samples with genotype A/A, 233 samples with genotype A/G, and 121 samples with genotype G/G.The median expression level for A/G is 25.065, for G/G is 20.074, and for A/A is 31.25


Make box plot
```{r}
library(ggplot2)
```
```{r}
ggplot(expr, aes(geno, exp, fill = geno)) +
  geom_boxplot(notch = T)
```
> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

Individuals with genotype A/A tend to have higher expression of ORMDL3. It seems like SNP effect the expression of this gene. 
