---
title: "Lab17_report"
author: "Longmei_Zhang_A17012012"
format: pdf
---

## Importing and filter read abundance results

Use tximport to import of Kallisto results
```{r}
library(tximport)

# setup the folder and filenames to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

txi.kallisto has estimated transcript counts for each sample in R. 
```{r}
head(txi.kallisto$counts)
```

Check the total number of transcripts for each sample:

```{r}
colSums(txi.kallisto$counts)
```
how many transcripts are detected in at least one sample. Number of genes with transcripts in at least one sample. 
```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

Before subsequent analysis, we might want to filter out those transcripts with no reads:
```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

And those with no change over the samples:
```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

## PCA

Now we compute the principal components, centering and scaling each transcript’s measured levels so that each feature contributes equally to the PCA:
```{r}
pca <- prcomp(t(x), scale=TRUE)
```

```{r}
summary(pca)
```

use the first two principal components as a co-ordinate system for visualizing the summarized transcriptomic profiles of each sample:

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```
> Q. Use ggplot to make a similar figure of PC1 vs PC2 and a seperate figure PC1 vs PC3 and PC2 vs PC3.

```{r}
head(x)
```

```{r}
library(ggplot2)
library(ggrepel)

# Make metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

# Make the data.frame for ggplot 
y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)

ggplot(y) +
  aes(PC1, PC2, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

```{r}

#PC1 vs PC3
ggplot(y) +
  aes(PC1, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

```{r}

##PC2 vs PC3
ggplot(y) +
  aes(PC2, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

## Differential Analysis

```{r}
library(DESeq2)
```

```{r}
sampleTable <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(sampleTable) <- colnames(txi.kallisto$counts)
```

```{r}

dds <- DESeq2::DESeqDataSetFromTximport(txi.kallisto, sampleTable, ~condition)

```

```{r}
dds <- DESeq2::DESeq(dds)
```

```{r}
res <- DESeq2::results(dds)
head(res)
```

