---
title: "class10/Halloween Candy"
author: "Longmei Zhang A17012012"
format: pdf
---

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
There are 85 different candy types in this dataset

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 fruity candy types in the dataset

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Milky Way", "winpercent"]
```
```{r}
library(dplyr)

candy|>
  filter(rownames(candy) == "Haribo Happy Cola") |>
  select(winpercent)
```

My favourite is MilkyWay, and its winpercent is 73.10%

> Q find candy with winpercent above 50%

```{r}
candy |>
  filter(winpercent > 50) |>
  filter(fruity == 1)

#same results
#top.candy <- candy[candy$winpercent > 50]
#top.candy[candy$fruity == 1]
```


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", "winpercent"]
```
The winepercent value of Kit kat is 76.78%

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", "winpercent"]
```
The winpercent value of Tootsie Roll Snack Bars is 49.65

To get a quick insight into a new dataset some folks like using skimr package and function `skim()`

```{r}
library("skimr")
skim(candy)
```


> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent value is measured on different scale than everything else. Need to scale the data before doing analysis like PCA. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

1 Means the candy is chocolate type while 0 means the candy is not chocolate type. 


> Q8. Plot a histogram of winpercent values

```{r}
#hist(candy$winpercent, breaks = 10)

library("ggplot2")
ggplot(candy, aes(candy$winpercent)) +
  geom_histogram(binwidth = 6, color = "dark grey") +
  theme_bw()
```

> Q9. Is the distribution of winpercent values symmetrical?

No, its skewed to the left

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```
Using median to represent the center, the center is blow 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
fruit.candy <- candy |> 
  filter(candy$fruity == 1)

summary(fruit.candy$winpercent)

choco.candy <- candy |> 
  filter(candy$chocolate == 1)

summary(choco.candy$winpercent)
```
Chocolate candy have a higher median and mean, meaning that chocolate candies are ranked higher on average. 

> Q12. Is this difference statistically significant?

```{r}
t.test(choco.candy$winpercent, fruit.candy$winpercent)
```
According to the p-value, which is very small, it is very unlikely to get this result by chance. The difference between winpercent of chocolate and fruity is significant. 

> Q13. What are the five least liked candy types in this set?

```{r}
#play <- c("d", "a", "c")
#sort(play)
#order(play)
head(candy[order(candy$winpercent), ], 5)
```

the least favourite candies are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing = T), ], 5)
```


> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy)+
  aes(winpercent, rownames(candy), fill = chocolate) +
  geom_col() +
  theme_classic()
```


> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy)+
  aes(winpercent, reorder(rownames(candy), winpercent), fill = chocolate) +
  geom_col() +
  theme_classic()
```
Designing a more specialized colore scheme where we can see both chocolate and bar and fruity etc. all from the same plot. Change the color vector. 

```{r}
mycols <- rep("black", nrow(candy))
mycols[as.logical(candy$bar)] = "brown"
mycols[as.logical(candy$chocolate)] = "chocolate"
mycols[as.logical(candy$fruity)] = "pink"

# Use blue for favorite candy
mycols[rownames(candy) == "Milky Way"] = "blue"
```
```{r}
ggplot(candy)+
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill = mycols)
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets
```{r}
choco = candy[as.logical(candy$chocolate), ]
head(choco[order(choco$winpercent), ],5)


```


> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst. 

```{r}
fruity = candy[as.logical(candy$fruity), ]
head(fruity[order(fruity$winpercent, decreasing = T), ],5)

```

##Taking a look at price percent


```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label = rownames(candy)) +
  geom_point(col = mycols) +
  geom_text(col = mycols)
```
Making labels visible and not overlapping
```{r}
library("ggrepel")
ggplot(candy) +
  aes(winpercent, pricepercent, label = rownames(candy)) +
  geom_point(col = mycols) +
  geom_text_repel(col = mycols, max.overlaps = 8, size = 3.3)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures is ranked high in terms of winpercent for the least money. 


> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The top 5 most expensive candies are Nik L Nip, Ring pop, Nestle Smarties, Hershey's Krackel, and Hershey's Milk Chocolate. Nik L Nip is the least popular. 

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```



## Exploring the correlation structure
```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij, diag = F)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and Fruity are most anti-correlated

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent, and chocolate and bar are most positively correlated. 

## PCA

```{r}
pca <- prcomp(candy, scale = T)
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2], col = mycols, pch=16)
```

We can make a much nicer plot with the ggplot2 package. ggplot works best when we supply an input data.frame that includes a separate column for each of the aesthetics you would like displayed in your final plot. To accomplish this we make a new data.frame here that contains our PCA results with all the rest of our candy data. We will then use this for making plots below

```{r}
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=mycols)

p
```

We can use the `ggrepel` package and the function `ggrepel::geom_text_repel()` to label up the plot with non overlapping candy names like


```{r}
p + geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

We can also generate interactive plot using `plotly`

```{r}

#library(plotly)
#ggplotly(p)
```


How do the original variables (columns) contribute to the new PCs. Looking at PC1

```{r}
loadings <- as.data.frame(pca$rotation)
ggplot(loadings) +
  aes(PC1, reorder(rownames(loadings), PC1)) +
  geom_col()
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity strongly contributed to PC1 in the positive direction. It make sense since Fruity candy types are mainly located on the right side (positive direction of PC1) of the PC1 vs. PC2 graph. 